---
title: "SchistoTransmissionModel: modelling schistosomiasis transmission & immunity"
author: "Gregoy Milne"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SchistoTransmissionModel: modelling schistosomiasis transmission}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Package set-up

The package can be installed and loaded like so:
```{r}
if (!require(SchistoTransmissionModel)) { install.packages("SchistoTransmissionModel", dependencies = TRUE)
}
```

## Model parameter definition

The function to run the transmission model is `RunTransmissionModel()`, which takes the following arguments:

- `theta`:  a vector of transmission and immunity relevant parameters
- `tx_pars`: a vector of treatment-relevant parameters
- `runtime`: the length of time (in years) to run the model for
- `stepsize`: the integration time step (for instance, 1/12 will give a timestep of 1 month, 1/52 a timestep of 1 week)
- `user_tx_times`: a vector of user-inputted MDA times (used to specify irregular intervals of MDA). Only used if `tx_pars["input_tx_times"]==1`; otherwise (if `tx_pars["input_tx_times"]==0`), parameters specified in `tx_pars` are used calculate MDA times.
- `user_cov_weight`: a vector of coverage weights to specify relative distribution of treatment coverage between school-aged children (SAC) vs. adults (see `cov_weight` below for further details). Used if `tx_pars["input_tx_times"]==1`. NB must be the same length as `user_tx_times`.
- `time_extract_states` a model time to extract the state matrices for females and males. This parameter is particularly useful if the user would like this model output to feed into the input of another model. If you don't need to use this parameter, it can safely be set to `NA`.

The transmission and immunity model parameters (the `theta` argument of `RunTransmissionModel()`) are defined (with example values) like so:
```{r}
# Define model parameters (names must be spelt exactly but order doesn't matter)
theta <- c(    
  R0                  = 2.5,
  R0_weight           = 0.4,
  NhNs                = 0.3,
  kW                  = 0.4,
  decay_immunity      = 6,
  protection_immunity = 0.5,
  epg_constant        = 5.81,
  ige_constant        = 0.5 
)
```

where:

- `R0` is the basic reproduction number
- `R0_weight` controls the relative weighting of R0, from snails-to-human vs. humans-to-snails
- `NhNs` is the human-to-snail population density
- `kW` is the among-host worm overdispersion parameter (inversely related to aggregation)
- `protection_immunity` is the protection afforded by smTAL1-IgE antibodies against cercarial establishment (rage: 0,1)
- `epg_constant` is the no. eggs shed per fertile female schistosome
- `ige_constant` is a constant parameter that maps acquired immunity to antibody optical densities

The treatment relevant parameters (the `tx_pars` argument) are defined (with example values) like so:
```{r}
# Define treatment parameters (names must be spelt exactly but order doesn't matter)
tx_pars <- c(
  input_tx_times = 0,
  toggle_tx      = 1,
  start_tx       = 20,
  n_tx           = 3,
  freq_tx        = 1,
  sac_coverage   = 0.75,
  efficacy       = 0.95,
  min_tx_age     = 5,
  max_tx_age     = 30,
  cov_weight     = 0.8
)
```

where:

- `input_tx_times`: toggle parameter; when set to 0, treatment times are calculated from input parameters in the `tx_pars` vector; when set to 1 the model uses a user-inputted vector of treatment times
- `toggle_tx`: toggle parameter; when set to 0, the model is simulated without treatment; when set to 1 the model simulates treatment using the other treatment-relevant parameters (below)
- `start_tx`: model time to start treatment (used when `toggle_tx==1`) 
- `n_tx`: number of treatments (integer)
- `freq_tx`: frequency of treatments in years; e.g. freq_tx=1 effects annual MDA, 0.5 twice-annual, 0.25 four-times annual, etc. Note that treatment frequencies must concord with the user-inputted model stepsize (above)
- `sac_coverage`: MDA coverage in SAC (5-15 year olds; a proportion)
- `efficacy`: drug efficacy against adult worms (a proportion from 0 [0% efficacy] to 1 [100% efficacy])
- `min_tx_age`: minimum age of individuals that are treated during MDA (an integer)
- `max_tx_age`: maximum age of individuals that are treated during MDA (an integer)
- `cov_weight`: weighting parameter that determines coverage in adults (16+ year olds), where adult coverage = `sac_coverage` * `cov_weight` (NB pre-SAC coverage is hard-coded as 0 in the model and cannot be altered)

We then set model run time (in years) and model time interval (integration stepsize) by:

```{r}
runtime <- 30     #30 years
stepsize <- 1/12  #1 month
```


## Model simulation

Now that all arguments are defined for `RunTransmissionModel`, the model can be run like so: 

```{r}
sim <- RunTransmissionModel(
  theta    = theta,
  tx_pars  = tx_pars, 
  runtime  = runtime, 
  stepsize = stepsize, 
  user_tx_times   = NA, 
  user_cov_weight = NA, 
  time_extract_states = NA
)
```

Note the use of `user_tx_times = NA` and `user_cov_weight = NA` (since `tx_pars["input_tx_times"]==0`) and `time_extract_states = NA`. We'll come back to other parameterisations shortly.


### Visualising model output
The `RunTransmissionModel()` function returns a range of outputs (to explore these, run `View(sim)`). Outputs can be broadly be split into those that are time-dependent and and those that are age- and sex-dependent. We can visualise time-dependent outputs like so (for later ease of demonstrating another example, we will place the plotting code in a function):

```{r, fig.show='hold'}
plot_time_output <- function(sim){
  plot(sim$time, sim$worm_burden, 'l', xlab = "Time (years)", ylab = "Mean worm burden")
  plot(sim$time, sim$epg, 'l', xlab = "Time (years)", ylab = "Mean epg")
}
plot_time_output(sim)
```

We can visualise age- and sex-dependent outputs at a particular time point of the simulation like so:

```{r, fig.show='hold'}
# Specify time index to extract model output
time_index <- which(sim$time == tx_pars["start_tx"])

# Define a function to plot worm burdens & epgs at particular time index
plot_age_output <- function(sim, time_index) {
  
  # Plot worm burdens
  plot(sim$age, sim$worm_burden_age_female[time_index,], 'l', xlab = "Age (years)", ylab = "Worm burden")
  lines(sim$age, sim$worm_burden_age_male[time_index,], lty=2, col="red")
  
  # Plot epgs
  plot(sim$age, sim$epg_age_female[time_index,], 'l', xlab = "Age (years)", ylab = "epg")
  lines(sim$age, sim$epg_age_male[time_index,], lty=2, col="red")
  
}

# Call function to make plots
plot_age_output(sim, time_index)
```

Note here that the model output (`sim`) is stored in a matrix, such that there is one row per time point. In this example, we therefore specify that we want to extract the age-dependent output at the index corresponding to the time at which MDA begins (`tx_pars["start_tx"]`); although the user can select whatever time point they wish.


### Specification of irregular MDA treatment times

As mentioned earlier, we can specify irregular treatment times via an alternative model parameterisation like so:

```{r}
# set toggle to 1 so model uses 'user_tx_times' and `user_cov_weight`, not MDA times and coverage weights informed by 'tx_pars' parameters
tx_pars["input_tx_times"] <- 1

# then specify some example treatment times
user_tx_times <- c(tx_pars["start_tx"], tx_pars["start_tx"]+5, tx_pars["start_tx"]+7)

# and associated coverage weights
user_cov_weight <- c(0.5, 1, 0.2)
```

We can also choose to extract the model state matrices for female and male hosts by specifying a time to do so, for instance:

```{r}
user_t_extract_states <- tx_pars["start_tx"] + 3  #3 years after first treatment
```

The model can then be simulated with this alternate parameterisation,

```{r}
sim <- RunTransmissionModel(
  theta    = theta,
  tx_pars  = tx_pars, 
  runtime  = runtime, 
  stepsize = stepsize, 
  user_tx_times   = user_tx_times, 
  user_cov_weight = user_cov_weight, 
  time_extract_states = user_t_extract_states
)
```

and time-dependent output visualised to demonstrate the effect on MDA timing,

```{r}
plot_time_output(sim)
```

We can also inspect the sex-specific state matrices at the time we chose to extract them from the model, like so,

```{r}
## Uncomment to run
# head(sim$male_states_out)
# head(sim$female_states_out)
```
